Transactional Memory Concurrency Verification with Landslide

### 2018-03-29 htm/for-garth/paper.tex
add for-garth version

%\documentclass[preprint]{sigplanconf}
\documentclass[10pt]{sigplanconf}

@@

\usepackage{yfonts}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amssymb}
%\usepackage{mathpartir}
\usepackage{hyperref}
\usepackage{url}
\usepackage{graphics}
\usepackage{graphicx}
\usepackage{wasysym}
\usepackage{harmony}
\usepackage{marvosym}
\usepackage{multirow}
\usepackage[usenames,dvipsnames]{xcolor}
\usepackage[charter]{mathdesign}
\usepackage{natbib}
\usepackage{xspace}
\usepackage[mathcal]{euscript}
\usepackage[linesnumbered,ruled]{algorithm2e}

@@

\conferenceinfo{SIGBOVIK '18}{Pittsburgh, PA, USA}
\copyrightyear{2018}
\copyrightdata{}

\titlebanner{banner above paper title}        % These are ignored unless
\preprintfooter{short description of paper}   % 'preprint' option specified.

@@

\authorinfo{Ben Blum}{}{bblum@cs.cmu.edu}

@@

\begin{abstract}
Hardware transactional memory is a recently-introduced concurrent programming paradigm
%backed by a set of new instructions on recent x86 CPUs,
which allows programmers to elide locks for performance in low-contention workloads.
However,
%using this feature
it
comes at a cost in implementation complexity:
%programmers must support their
fast-path code must be accompanied by backup paths to handle transaction failure.
%
We extend Landslide, a popular stateless model checker,
with a concurrency model for transactional memory
and evaluate it on several real-world transactional
benchmarks and data structure implementations.
\end{abstract}

\category{D.1.3}{Programming Techniques}{Concurrent Programming}
\category{D.2.4}{Software Engineering}{Software/Program Verification}

\keywords
landslide terminal, baggage claim, ground transportation, ticketing

\section{Introduction}
\label{sec:intro}

Transactional Synchronization Extensions (TSX) \cite{transactional-memory}
is an instruction set extension for x86 CPUs which adds hardware-based transactional memory.
The processor uses its existing cache coherence algorithm to
check for memory conflicts with other cores while temporarily staging a sequence of memory accesses.
If no other CPU accesses the same memory during the transaction,
the access sequence is committed to main memory atomically (with respect to visibility by other CPUs).
Otherwise, the accesses are discarded, the CPU's local state is reverted, and the transaction returns a failure code.

This feature can be used to replace conventional locking in performance-critical concurrent programs.
When a concurrent workload accesses largely thread-local data,
or disjoint sections of a shared data structure,
the contention rate between threads is low,
and transactions will often succeed.
Compared to programs which use conventional locks,
which use bus-locking atomic accesses even in the fastest code path,
TSX provides substantial performance improvements in such programs \cite{htm-experience, htm-performance, htm-mario}.
However, the possibility of transaction failure introduces additional implementation complexity:
programmers must also provide a backup plan
to safely resolve contention between threads,
usually involving conventional synchronization.
These backup paths must coordinate not only with other backup paths
but also with other fast paths which another thread may begin after the original transaction failed,
which even in the simplest transactions requires complex synchronization sequences
\cite{htm-mario}.
This introduces an additional dimension of nondeterminism into an already concurrent program,
and moreover, because transaction failure is expected to be rare,
% だからこそ
obscure interleavings between failure paths are difficult to expose during stress testing.

This motivates the use of stateless model checking (MC) \cite{verisoft}
to comprehensively verify these transactional programs,
fast paths failure paths and all.
MC aims to force the system to execute all possible thread interleavings under a given test case,
exhaustively checking for bugs or verifying their absence in the corresponding state space.
Many such model checkers exist, varying in
interleaving granularity, memory analysis, types of programs checked, and search ordering strategy
\cite{chess,dbug-ssv,spin,inspect,r4,portend,samc,mcr,randomized-scheduler}.
This work builds upon Landslide \cite{landslide}, %,quicksand,landslide-phdthesis},
a simulator-based tester which checks both user- and kernel-level programs
and incorporates data-race analysis \cite{tsan,fasttrack} to find new preemption points
at memory access granularity.
Our contributions are as follows:

\begin{enumerate}
	\item We extend Landslide's concurrency model to include transaction failure as an additional source of nondeterminism;
	\item We provide a proof sketch that our implementation matches TSX's execution semantics,
	\item We evaluate the extended Landslide on several transactional programs, analyzing both its bug-finding and verification performance.
\end{enumerate}

The paper is organized as follows.
Section \ref{sec:intro} introduces the problem domain and motivates our research.
The other sections state the rest of the paper.

@@

This section introduces the fundamental concepts and prior work
in both hardware transactional memory and stateless model checking,
which we propose to combine.

{\bf Hardware transactional memory.}
TSX was implemented on consumer hardware for the first time by Intel's Haswell architecture \cite{htm-haswell},
which extends the x86 instruction set to provide
{\tt xbegin}, {\tt xend}, and {\tt xabort} for beginning, committing, and aborting
transactions, respectively.
Higher-level programming languages or compilers may offer libraries or intrinsics to access these instructions;
for C and C++ GCC provides intrinsics named {\tt \_xbegin()} and so on \cite{htm-gcc}.
Figure~\ref{fig:htm-example} shows an example program using TSX to synchronize access to a shared counter,
including a failure path which defaults to a conventional lock.
This example actually has a bug, which we will discuss in the next section;
the reader is encouraged to try to spot it before then.
Several related works formally prove the correctness of transactional memory {\em implementations}
\cite{specifying-verifying-tm,tm-correctness,tm-completeness,mc-tm-with-spin},
but verifying the client programs written to use transactions remains an open problem.

@@

Under software transactional memory (STM) \cite{stm-pldi06},
memory conflicts with other threads are the only reason for transaction failure
(apart from programmer-supplied explicit aborts);
hence, depending on program semantics, some transactions may be guaranteed to succeed.
However, hardware transactions (HTM) may also fail
for several other reasons such as random system interrupts or exhausting the CPU's cache capacity.
Because timer interrupts can in principle occur at any moment,
and with arbitrary frequency (observable by the program, perhaps as a result of a heavily-loaded system),
in this paper we will simplify the failure model by saying that HTM transactions can fail for any reason.
We defer discussion of programs which distinguish the reason for aborts through the failure code to Section~\ref{sec:warpzone}.

{\bf Stateless model checking.}
Model checking (MC) \cite{verisoft} is a testing technique for systematically executing and verifying
the possible thread interleavings of a concurrent program.
The main research challenge is to cope with exponential explosion of the state space,
which is sized $O(n^k)$ for a program with $n$ operations and $k$ threads.
Some {\em stateful} MCs explicitly store and compare all visited states of the program being tested \cite{spin},
which both keeps track of test coverage
and allows identifying identical states to avoid testing redundant interleavings.
%
By contrast, {\em stateless} MC (henceforth abbreviated simply as MC)
stores only the current sequence of execution events to avoid a prohibitive memory footprint.
Reduction algorithms \cite{dpor,optimal-dpor,satcheck,mcr,mcr,tsopso}
can then analyze the memory accesses in that sequence to identify
interleavings observationally-equivalent under Mazurkiewicz trace theory \cite{mazurkiewicz} and hence safe to skip.
The resulting state spaces are still exponentially-sized,
but only in the number of conflicting operations rather than all operations.
Of these, Landslide uses Dynamic Partial Order Reduction (DPOR) \cite{dpor} to prune its state spaces.

\begin{figure}[t]
	\begin{center}
		\begin{tabular}{ll}
		%\texttt{void count() \{} \\
		%\texttt{~~~~for (int i = 0; i < 1000; i++) \{} \\
		1 & \texttt{\flow{if} ((status = \call{\_xbegin}()) == \const{SUCCESS}) \{} \\
		2 & \texttt{~~~~x++;} \\
		3 & \texttt{~~~~\call{\_xend}();} \\
		4 & \texttt{\} \flow{else} \{} \\
		5 & \texttt{~~~~\call{mutex\_lock}(\&m);} \\
		6 & \texttt{~~~~x++;} \\
		7 & \texttt{~~~~\call{mutex\_unlock}(\&m);} \\
		8 & \texttt{\}} \\
		%\texttt{~~~~\}} \\
		%\texttt{\}} \\
		\end{tabular}
	\end{center}
	\caption{Example transactional program.
		%The example {\tt count} routine from Figure~\ref{fig:mutex}, rewritten to use HTM.
		If the
		%transaction in the
		top branch aborts,
		%whether from a memory conflict or random system interrupt,
		%from the programmer's intention,
		execution will revert to the return of {\tt \_xbegin()}
		%{\tt status} will be assigned an error code
		and control will drop into the {\tt else} branch.
		The programmer can then use explicit synchronization
		%, such as a mutex,
		to resolve the conflict.}
	\label{fig:htm-example}
\end{figure}

MCs may instrument programs to introduce thread switches at varying granularity, which affects the number of operations $n$.
%The number of operations depends on how finely-grained the MC instruments the program to introduce thread switches.
Some target distributed systems, instrumenting only message-passing events \cite{modist};
some run multithreaded programs natively, instrumenting only the pthread API for performance \cite{dbug-ssv};
and some insert compiler instrumentation on statically-identified memory accesses \cite{chess,inspect}.
Landslide traces every memory access through the use of a simulated environment \cite{bochs},
which is important for identifying data races to use as new preemption points \cite{quicksand},
as well as for identifying when a memory conflict may cause transaction aborts.
With regard to checking for bugs,
the ``model'' the name refers to being checked may be
an external formal specification,
the program's own internal consistency checks,
or a set of expected properties encoded in the tool itself.
Landslide uses the latter two cases, checking for assertion failures
as well as deadlocks, use-after-frees, and segfaults.
For this work we also detect use of {\tt xend} outside of a transaction
or {\tt xbegin} within one as a bug.

\section{Design}
\label{sec:design}

This section presents our formalization of transactional memory in Landslide's framework of thread concurrency.
We make two major simplifications:
simulating transaction aborts as immediate failure injections,
and treating transaction atomicity as a global mutex during data-race analysis;
and provide corresponding equivalence proofs.

{\bf Notation.} Let $I = TN_1@L_1, TN_2@L_2, ... TN_n@L_n$,
with $N_i$ a thread ID and $L_i$ a code line number,
denote the execution sequence of a program as it runs according to the specified thread interleaving.%
\footnote{This serialization of concurrent execution is told from the perspective of all CPUs at once
and hence assumes sequential consistency.
For discussion of relaxed memory models refer to Section~\ref{sec:warpzone}.}

\subsection{Example}

Consider again the program in Figure~\ref{fig:htm-example}.
Note that the C-style {\tt x++} operations, when compiled into assembly \cite{printy},
results in multiple memory accesses which can be interleaved with other threads.
\[
\begin{tabular}{ll}
	$2a$ & \texttt{temp <- x;} \\
	$2b$ & \texttt{temp <- temp + 1;} \\
	$2c$ & \texttt{x <- temp;} \\
\end{tabular}
\]

% T1 T2 T3 but latex doesn't allow #s in cmd names
\newcommand\ti{\ensuremath{\hilight{lavender}{\mathbf{T1}}}\xspace}
\newcommand\tj{\ensuremath{\hilight{seafoam}{\mathbf{T2}}}\xspace}
\newcommand\tk{\ensuremath{\hilight{salmon}{\mathbf{T3}}}\xspace}

If these instructions from the {\tt x++} in the transaction are preempted,
with another thread's access to {\tt x} interleaved in between,
the transaction will abort.
So, the interleaving
\[
	\tiat{1}, \tiat{2a}, \tiat{2b}, \tjat{1}, \tjat{2}, \tjat{3}, \tiat{2c}, \tiat{3}
\]
or, henceforth abbreviated for clarity:
\[
	\tiat{1-2b}, \tjat{1-3}, \tiat{2c-3}
\]
is not possible; rather, \ti will fall into the backup path:
\[
	\tiat{1-2b}, \tjat{1-3}, \tiat{4-7}
\]
However, the {\tt x++} operation from the failure path (correspondingly $6a$, $6b$, $6c$)
{\em can} be thusly separated with conflicting accesses interleaved in between,
since the mutex only protects the failure path against other failure paths,
but not against the transaction itself.
So (assuming {\tt x} is intended to be a precise counter rather than a sloppy one),
%losing one of the increments to which constitutes a bug),
we observe a bug in the following interleaving.%
\footnote{Note also that this bug requires either at least 3 threads or at least 2 iterations between 2 threads to expose;
this highlights MC's dependence on its test cases to produce meaningful state spaces in the first place.}
\[
	\tiat{1-2b}, \tjat{1-3}, \tiat{4-6b}, \tkat{1-3}, \tiat{6c-7}
\]
Prior work \cite{htm-mario} proposed the idiom shown in Figure~\ref{fig:htm-fixed}
to exclude this family of interleavings,
which shows that correctly synchronizing even the simplest transactions may be surprisingly difficult or complex.
%further motivating our research.

\begin{figure}[t]
	\begin{center}
		\begin{tabular}{ll}
		%\texttt{void count() \{} \\
		%\texttt{~~~~for (int i = 0; i < 1000; i++) \{} \\
		  & \texttt{prevent\_transactions = \const{false};} \\
		\\
		0 & \texttt{\flow{while} (prevent\_transactions) \flow{continue};} \\
		1 & \texttt{\flow{if} ((status = \call{\_xbegin}()) == \const{SUCCESS}) \{} \\
		2 & \texttt{~~~~\flow{if} (prevent\_transactions)} \\
		3 & \texttt{~~~~~~~~\call{\_xabort}();} \\
		4 & \texttt{~~~~x++;} \\
		5 & \texttt{~~~~\call{\_xend}();} \\
		6 & \texttt{\} \flow{else} \{} \\
		7 & \texttt{~~~~\call{mutex\_lock}(\&m);} \\
		8 & \texttt{~~~~prevent\_transactions = \const{true};} \\
		9 & \texttt{~~~~x++;} \\
		A & \texttt{~~~~prevent\_transactions = \const{false};} \\
		B & \texttt{~~~~\call{mutex\_unlock}(\&m);} \\
		C & \texttt{\}} \\
		%\texttt{~~~~\}} \\
		%\texttt{\}} \\
		\end{tabular}
	\end{center}
	\caption{Variant of the program in Figure~\ref{fig:htm-example},
		with additional synchronization to protect the failure path from the transactional path.
		The optional line 0 serves to prevent a cascade of failure paths
		for the sake of performance
		by allowing threads to wait until transacting is safe again.}
	\label{fig:htm-fixed}
\end{figure}

Left unstated in interleavings such as $\tiat{1-2c}, \tjat{1-3}, \tiat{4-7}$%
\footnote{For a clearer example we reorder \ti's write to {\tt x} before \tj's part here.}
are HTM's execution semantics, namely:
\begin{enumerate}
	\item any modifications to shared state (such as $2c$) by \ti are not visible to \tj during its execution,
		despite \tj being executed afterwards, and
	\item all local and global state changes by \ti between lines $1$ and $2c$ are discarded when jumping to line $4$.
\end{enumerate}
% there shouldbe a better word than 'requires' here.. benefits from.. "the point of it is"..
While use of TSX in production requires the performance advantage
of temporarily staging such accesses in local CPU cache,
model checking such programs need be concerned only with the program's {\em observable} behaviours.
We claim that MCing the simpler interleaving $\tiat{1}, \tjat{1-3}, \tiat{4-7}$
is an equivalent verification as MCing the one above;
in fact, this interleaving suffices to check
all observable behaviours
of all interleavings
of all subsets of $\tjat{1-3}$
with all subsets of $\tiat{2a-2c}$,
whether they share a memory conflict or not.
Stated formally:

@@

\newcommand\llama[1]{\includegraphics[width=#1]{llama.pdf}}
\newcommand\llitem{\item[\raisebox{-0.15em}{\llama{1.2em}}]}

\begin{lemma}[Equivalence of Aborts]
	\label{lem:equiv}
	Let:
	\begin{itemize}
		\item $\tiiat{\alpha}$ be an HTM begin operation,
		\item $\tiiat{\beta_1}\dots\tiiat{\beta_n}$ be the transaction body (with $\beta_n$ the HTM end call),
		\item $\tiiat{\phi_1}\dots\tiiat{\phi_m}$ be the failure path, and
		\item $\tiiat{\omega_1}\dots\tiiat{\omega_l}$ be the subsequent code executed unconditionally.%
			\footnote{Arbitrary code may not be structured to distinguish these as nicely as in our examples;
			e.g., more code may exist in the success branch after {\tt \_xend()};
			such would be considered part of $\omega$ here.}
	\end{itemize}
	Then, for any interleaving prefix%
	\footnote{Without loss of generality: for any number of other threads \tjj/\tkk,
	and for any number of thread switches away from \tii during the transaction.}
	\[
	\begin{tabular}{c}
		$\tiiat{\alpha},\tiiat{\beta_1}\dots\tiiat{\beta_b},$\\
		$\tjjat{\gamma_1}\dots\tjjat{\gamma_j},$ \\
		$\tkkat{\kappa_1}\dots\tkkat{\kappa_k},$ \\
		$\tiiat{\beta_{b+1}}$ % $\dots\tiiat{\beta_{n-1}}$ -- excluded bc might abort
	\end{tabular}
	\]
	with $b<n$, $j \ne i$, $k \ne i$, etc., either:
	\begin{enumerate}
		\item $\tiiat{\alpha},\tjjat{\gamma_1}\dots\tjjat{\gamma_j},\tkkat{\kappa_1}\dots\tkkat{\kappa_k},\tiiat{\phi_1}\dots$
			(conflicting case), or
		\item $\tiiat{\alpha},\tiiat{\beta_1}\dots\tiiat{\beta_b}\dots\tiiat{\beta_n},\tjjat{\gamma_1}\dots\tjjat{\gamma_j},$ \\
			$\tkkat{\kappa_1}\dots\tkkat{\kappa_k}$
			(independent case)
	\end{enumerate}
	exists and is observationally equivalent.
\end{lemma}

>>>>

%\documentclass[preprint]{sigplanconf}
\documentclass{article}

@@

\usepackage{yfonts}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amssymb}
%\usepackage{mathpartir}
\usepackage{hyperref}
\usepackage{url}
\usepackage{graphics}
\usepackage{graphicx}
\usepackage{wasysym}
\usepackage{harmony}
\usepackage{marvosym}
\usepackage{multirow}
\usepackage[usenames,dvipsnames]{xcolor}
\usepackage[charter]{mathdesign}
\usepackage{natbib}
\usepackage{xspace}
\usepackage{fullpage}
\usepackage[mathcal]{euscript}
\usepackage[linesnumbered,ruled]{algorithm2e}

@@

%\conferenceinfo{SIGBOVIK '18}{Pittsburgh, PA, USA}
%\copyrightyear{2018}
%\copyrightdata{}
%
%\titlebanner{banner above paper title}        % These are ignored unless
%\preprintfooter{short description of paper}   % 'preprint' option specified.

@@

\author{Ben Blum \\ bblum@cs.cmu.edu}
\date{}

@@

%\category{D.1.3}{Programming Techniques}{Concurrent Programming}
%\category{D.2.4}{Software Engineering}{Software/Program Verification}
%
%\keywords
%landslide terminal, baggage claim, ground transportation, ticketing

@@

\subsection{Hardware transactional memory}

Transactional Synchronization Extensions (TSX) was implemented on consumer hardware for the first time by Intel's Haswell architecture \cite{htm-haswell},
which extends the x86 instruction set to provide
{\tt xbegin}, {\tt xend}, and {\tt xabort} for beginning, committing, and aborting
transactions, respectively.
Higher-level programming languages or compilers may offer libraries or intrinsics to access these instructions;
for C and C++ GCC provides intrinsics named {\tt \_xbegin()} and so on \cite{htm-gcc}.
Figure~\ref{fig:htm-example} shows an example program using TSX to synchronize access to a shared counter,
including a failure path which defaults to a conventional lock.
%This example actually has a bug, which we will discuss in the next section;
%the reader is encouraged to try to spot it before then.
Several related works formally prove the correctness of transactional memory {\em implementations}
\cite{specifying-verifying-tm,tm-correctness,tm-completeness,mc-tm-with-spin},
but verifying the client programs written to use transactions remains an open problem.

@@

Under software transactional memory (STM) \cite{stm-pldi06},
memory conflicts with other threads are the only reason for transaction failure
(apart from programmer-supplied explicit aborts);
hence, depending on program semantics, some transactions may be guaranteed to succeed.
However, hardware transactions (HTM) may also fail
for several other reasons such as random system interrupts or exhausting the CPU's cache capacity.
Because timer interrupts can in principle occur at any moment,
and with arbitrary frequency (observable by the program, perhaps as a result of a heavily-loaded system),
in this paper we will simplify the failure model by saying that HTM transactions can fail for any reason.
%We defer discussion of programs which distinguish the reason for aborts through the failure code to Section~\ref{sec:warpzone}.

\subsection{Stateless model checking}

Model checking (MC) \cite{verisoft} is a testing technique for systematically executing and verifying
the possible thread interleavings of a concurrent program.
The main research challenge is to cope with exponential explosion of the state space,
which is sized $O(n^k)$ for a program with $n$ operations and $k$ threads.
Some {\em stateful} MCs explicitly store and compare all visited states of the program being tested \cite{spin},
which both keeps track of test coverage
and allows identifying identical states to avoid testing redundant interleavings.
%
By contrast, {\em stateless} MC (henceforth abbreviated simply as MC)
stores only the current sequence of execution events to avoid a prohibitive memory footprint.
Reduction algorithms \cite{dpor,optimal-dpor,satcheck,mcr,mcr,tsopso}
can then analyze the memory accesses in that sequence to identify
interleavings observationally-equivalent under Mazurkiewicz trace theory \cite{mazurkiewicz} and hence safe to skip.
The resulting state spaces are still exponentially-sized,
but only in the number of conflicting operations rather than all operations.
%Of these, Landslide uses Dynamic Partial Order Reduction (DPOR) \cite{dpor} to prune its state spaces.

\begin{figure}[t]
	\begin{center}
		\begin{tabular}{ll}
		%\texttt{void count() \{} \\
		%\texttt{~~~~for (int i = 0; i < 1000; i++) \{} \\
		1 & \texttt{\flow{if} (\call{\_xbegin}() == \const{SUCCESS}) \{} \\
		2 & \texttt{~~~~x++;} \\
		3 & \texttt{~~~~\call{\_xend}();} \\
		4 & \texttt{\} \flow{else} \{} \\
		5 & \texttt{~~~~\call{mutex\_lock}(\&m);} \\
		6 & \texttt{~~~~x++;} \\
		7 & \texttt{~~~~\call{mutex\_unlock}(\&m);} \\
		8 & \texttt{\}} \\
		%\texttt{~~~~\}} \\
		%\texttt{\}} \\
		\end{tabular}
	\end{center}
	\caption{Example transactional programs.
			If the
			%transaction in the
			top branch aborts,
			%whether from a memory conflict or random system interrupt,
			%from the programmer's intention,
			execution will revert to the return of {\tt \_xbegin()}
			%{\tt status} will be assigned an error code
			and control will drop into the {\tt else} branch.
			The programmer can then use explicit synchronization
			%, such as a mutex,
			to resolve the conflict.
		}
	\label{fig:htm-example}
\end{figure}
MCs may instrument programs to introduce thread switches at varying granularity, which affects the number of operations $n$.
%The number of operations depends on how finely-grained the MC instruments the program to introduce thread switches.
Some target distributed systems, instrumenting only message-passing events \cite{modist};
some run multithreaded programs natively, instrumenting only the pthread API for performance \cite{dbug-ssv};
and some insert compiler instrumentation on statically-identified memory accesses \cite{chess,inspect}.
Our model checker, Landslide \cite{landslide},
traces every memory access through the use of a simulated environment \cite{bochs},
which is important for identifying data races to use as new preemption points \cite{quicksand},
as well as for identifying when a memory conflict may cause transaction aborts.
With regard to checking for bugs,
the ``model'' the name refers to being checked may be
an external formal specification,
the program's own internal consistency checks,
or a set of expected properties encoded in the tool itself.
Landslide uses the latter two cases, checking for assertion failures
as well as deadlocks, use-after-frees, and segfaults.
%For this work we also detect use of {\tt xend} outside of a transaction
%or {\tt xbegin} within one as a bug.

\subsection{Example bug}
\label{sec:example}

@@

Consider again the program in Figure~\ref{fig:htm-example}.
Note that the C-style {\tt x++} operations, when compiled into assembly.
results in multiple memory accesses which can be interleaved with other threads.
\[
\begin{tabular}{ll}
	$2a$ & \texttt{temp <- x;} \\
	$2b$ & \texttt{temp <- temp + 1;} \\
	$2c$ & \texttt{x <- temp;} \\
\end{tabular}
\]

The {\tt x++} operation from the failure path
%(correspondingly $6a$, $6b$, $6c$)
can by interrupted with conflicting accesses from the transaction path interleaved in between,
since the mutex only protects the failure path against other failure paths,
but not against the transaction itself.
Hence an interleaving such as the following is possible,
in which only two increments of {\tt x} are observed while three are expected.

\[
\begin{tabular}{ll|ll|ll}
	\multicolumn{2}{c|}{\hilight{lavender}{Thread 1}} &
	\multicolumn{2}{c|}{\hilight{seafoam}{Thread 2}} &
	\multicolumn{2}{c}{\hilight{salmon}{Thread 3}} \\
	\hline
	\hilight{lavender}{$1$} & \hilight{lavender}{\texttt{\_xbegin();}} & & \\
	%
	& & \hilight{seafoam}{$1$} & \hilight{seafoam}{\texttt{\_xbegin();}} \\
	& & \hilight{seafoam}{$2$} & \hilight{seafoam}{\texttt{x++;}} \\
	& & \hilight{seafoam}{$3$} & \hilight{seafoam}{\texttt{\_xend();}} \\
	%
	\hilight{lavender}{$5$} & \hilight{lavender}{\texttt{mutex\_lock(m);}} & & \\
	\hilight{lavender}{$6a$} & \hilight{lavender}{\texttt{temp <- x;}} & & \\
	%
	& & & & \hilight{salmon}{$1$} & \hilight{salmon}{\texttt{\_xbegin();}} \\
	& & & & \hilight{salmon}{$2$} & \hilight{salmon}{\texttt{x++;}} \\
	& & & & \hilight{salmon}{$3$} & \hilight{salmon}{\texttt{\_xend();}} \\
	%
	\hilight{lavender}{$6b$} & \hilight{lavender}{\texttt{temp <- temp + 1;}} & & \\
	\hilight{lavender}{$6c$} & \hilight{lavender}{\texttt{x <- temp;}} & & \\
	\hilight{lavender}{$7$} & \hilight{lavender}{\texttt{mutex\_unlock(m);}} & & \\
\end{tabular}
\]

Prior work \cite{htm-mario} proposed the idiom shown in Figure~\ref{fig:htm-fixed}
to exclude interleavings which violate the failure path's atomicity.
Thread 1 would set the {\tt prevent\_transactions} flag across its {\tt x++} accesses,
and
Thread 3 would have to check it and trigger an abort. %(Figure \ref{fig:htm-fixed} lines 2-3).
%which shows that correctly synchronizing even the simplest transactions may be surprisingly difficult or complex.
%further motivating our research.

\begin{figure}[t]
	\begin{center}
		\begin{tabular}{ll}
			%\texttt{void count() \{} \\
			%\texttt{~~~~for (int i = 0; i < 1000; i++) \{} \\
			  & \texttt{prevent\_transactions = \const{false};} \\
			\\
			0 & \texttt{\flow{while} (prevent\_transactions) \flow{continue};} \\
			1 & \texttt{\flow{if} (\call{\_xbegin}() == \const{SUCCESS}) \{} \\
			2 & \texttt{~~~~\flow{if} (prevent\_transactions)} \\
			3 & \texttt{~~~~~~~~\call{\_xabort}();} \\
			4 & \texttt{~~~~x++;} \\
			5 & \texttt{~~~~\call{\_xend}();} \\
			6 & \texttt{\} \flow{else} \{} \\
			7 & \texttt{~~~~\call{mutex\_lock}(\&m);} \\
			8 & \texttt{~~~~prevent\_transactions = \const{true};} \\
			9 & \texttt{~~~~x++;} \\
			A & \texttt{~~~~prevent\_transactions = \const{false};} \\
			B & \texttt{~~~~\call{mutex\_unlock}(\&m);} \\
			C & \texttt{\}} \\
		\end{tabular}
	\end{center}
	\caption{Variant %of the program in Figure~\ref{fig:htm-example},
		with additional synchronization to protect the failure path from the transactional path.
		The optional line 0 serves to prevent a cascade of failure paths
		for the sake of performance
		by allowing threads to wait until transacting is safe again.
		}
	\label{fig:htm-fixed}
\end{figure}

\section{Implementation}

To model check a transactional program it suffices to test a set of interleavings
corresponding to all {\em observable} behaviours of the program.
Because incomplete transactions are rolled back upon abort,
we claim that interleaving other threads within the body of a transaction is never necessary,
and hence model checking can achieve full coverage simply by conisdering each {\tt xbegin} invocation
to be a preemption point with nondeterministic failure injection.
Appendix~\ref{sec:proofs} provides several proof sketches to justify this claim.

Our implementation of HTM-equivalent semantics has been incorporated by the Landslide maintainers upstream.
It is available open-source at \url{https://github.com/bblum/landslide}.
Programs should be ported to the Pebbles userland \cite{thrlib,kspec},
their use of compiler HTM intrinsics should be replaced with the Landslide stubs provided in {\tt 410user/inc/htm.h},
and HTM nondeterminism can then be enabled with the {\tt -X} command-line flag.
%We have also extended its Iterative Deepening implementation \cite{quicksand}
%with a new option ({\tt -M} flag) to
%to optimize for completion time by prioritizing the maximal state space job and cancelling all others
%(which maintains its soundness guarantee)
%which we will use in our evaluation.
%All test cases therein are also available in the repository linked above.
All test cases in our evaluation to follow are also available in the repository linked above.

%To the best of our knowledge, this is the first work to test transactional programs in a model-checking environment,
%%To the best of the author's knowledge, Landslide is the first MC to check transactional concurrency,
%so no other MC State of the Art%
%\footnote{The author's DJ name.}
%exists to compare to in controlled experiments.
%Nevertheless, we pose the following evaluation questions.

%\begin{enumerate}
%	\item How quickly does Landslide find bugs in incorrect transactional programs of varying sizes?
%	\item How quickly does Landslide verify correct transactional programs of varying sizes?
%	%\item Does Landslide find any previously-unknown bugs in real-world transactional code?
%	\item By the way, should MC research papers quantify variance in their CPU-time performance experiments?
%\end{enumerate}

Our preliminary evaluation evaluates Landslide's bug-finding performance on tests with subtle concurrency errors,
as well as its verification performance on (as far as we know) correct tests.
So far we have several hand-written unit tests
and \cite{htm-mario}'s microbenchmarks and transactional AVL tree and separate-chaining hashmap,
as follows.

\begin{itemize}
	\item {\tt htm1}: The bug from Figure~\ref{fig:htm-example}. %(assertion failure on {\tt count}).
	\item {\tt htm2}: The fixed version as in Figure~\ref{fig:htm-fixed}.
	\item {\tt counter}: Microbenchmark version of {\tt htm2} which replaces the complex failure path with a simple {\tt xadd}.
	\item {\tt swap}: Microbenchmark that swaps values in an array.
	\item {\tt swapbug}: {\tt swap} modified to introduce circular locking in the failure path. %(deadlock).
	\item {\tt avl\_insert}: AVL tree concurrent insertion test.
	\item {\tt avl\_fixed}: {\tt avl\_insert} with the AVL bug fixed. %(spoilers!!).
	\item {\tt map\_basic}: Separate-chaining hashmap concurrent insertion test.
	\item {\tt map\_basicer}: {\tt map\_basic} modified with a larger initial size to skip the resizing step.
\end{itemize}

Table~\ref{tab:buges}
presents our bug-finding results.
%We configured Landslide to run the Quicksand algorithm \cite{quicksand}
%% this note obsolete since swapbug was fixed and ends at 3,2
%%\footnote{This was necessary in {\tt swapbug}s (4,1) and (4,2) to trigger a switch to the buggy state space.}
%shown left,
%as well as to prioritize the maximal state space as discussed above,
%shown right,
%each with a time limit of 30 minutes.
%
Most notably,
in addition to the bugs we intentionally wrote in {\tt htm1} and {\tt swapbug},
to our pleasant surprise
Landslide also found a previously-unknown bug in the transactional AVL tree,
exposed by {\tt avl\_insert} with any parameters higher than (2,1).
Figure~\ref{fig:avlbug} shows the root cause, essentially the {\tt htm1} bug in disguise.
% landslide-trace-1521093835.18927.html
This manifested
alternately
%in (2,2) and (3,1)
as a segfault
% landslide-trace-1521092404.23772.html
and
%in (2,3)
as a consistency-check assertion failure.
The presence of {\tt while (\_retry);} makes the necessary preemption window extremely small
in practice (between it and {\tt \_xbegin()}),
whereas MC is blind to such matters of chance.%
% NB. In case a reader doesn't believe this footnote,
% I put the loop back in (with thr_yield(-1) therein) to restore the original state space,
% and found the bug with these statistics:
% [JOB 28] progress: 1957/3005897 brs (0.065105%), ETA 6414673y 121d 18h 28m 59s (elapsed 3m 15s)
% [JOB 28] BUG FOUND: landslide-trace-1521099044.76395.html (1958 interleavings tested; 3 preemptions; job time 3m 15s; pldi time 746832861; new-fixed pldi cputime 823243872)
%\footnote{Considering the loop does not affect the test's possible behaviours,
%only its {\em likely} ones,
%we removed it in Landslide's version of the test to keep the state space size manageable.}
%We removed the loop in Landslide's version of the test
%%To make this test Landslide-friendly, we removed that loop,
%which reduces the state space size, but maintains the program's semantics.}
The fact that even Figure~\ref{fig:htm-fixed}'s protocol's very proposer % こそさえ
didn't get it right every time %motivates the need for MC on such programs,
%and perhaps also means TSX primitives should be used only behind higher-level abstractions
%such as lock elision \cite{lock-elision}.
%and
suggests TSX primitives should be encapsulated behind higher-level abstractions
such as lock elision \cite{lock-elision},
which can be verified in isolation with smaller state spaces
then trusted in turn when checking their client programs \cite{dbug-phdthesis}.

\begin{table*}[h]
	\begin{tabular}{cc||r|r|r|r}
		\bf buggy test	& \bf params&\cpu{\bf cpu (s)}&\wtm{\bf wall (s)}&\ints{\bf int's}& \ETA{\bf SS size (est.)} \\
			& 2,1	&\cpu{*9.47}& \wtm{*6.40}& \ints{21	}& \ETA{213} \\
			& 2,2	&\cpu{*10.39}& \wtm{*7.70}& \ints{49	}& \ETA{1536} \\
			& 2,3	&\cpu{*12.83}& \wtm{*9.67}& \ints{113	}& \ETA{10752} \\
			& 2,4	&\cpu{*18.63}& \wtm{*15.86}& \ints{257	}& \ETA{73728} \\
		\cline{2-6}
			& 3,1	&\cpu{*9.50}& \wtm{*6.79}& \ints{21	}& \ETA{13653} \\
			& 3,2	&\cpu{*10.72}& \wtm{*7.97}& \ints{49	}& \ETA{393216} \\
			& 3,3	&\cpu{*13.84}& \wtm{*11.01}& \ints{113	}& \ETA{11010048} \\
			& 3,4	&\cpu{*31.37}& \wtm{*28.53}& \ints{257	}& \ETA{301989888} \\
		\cline{2-6}
			& 4,1	&\cpu{*9.82}& \wtm{*7.00}& \ints{21	}& \ETA{873813} \\
			& 4,2	&\cpu{*11.54}& \wtm{*8.55}& \ints{49	}& \ETA{100663296} \\
			& 2,1	&\cpu{*38.96}& \wtm{*13.15}& \ints{109	}& \ETA{194} \\
			& 2,2	&\cpu{*44.73}& \wtm{19.47}& \ints{281	}& \ETA{1620} \\
			& 2,3	&\cpu{*60.30}& \wtm{*35.55}& \ints{718	}& \ETA{12748} \\
			& 2,4	&\cpu{*108.58}& \wtm{82.60}& \ints{1820	}& \ETA{97823} \\
